Cartesian Closed Categories as a Model for the Simply Typed Lambda Calculus

### 2017-09-20 papers/talk-9/talk-9.tex
fixes

Recall the Curry-Howard correspondence which relates logic and type theory.
According to this correspondence proposotions are types and proofs are terms.
A prominent example is the relation between intuitionistic propositional logic and simply typed lambda calculus.
We can extend the correspondence to the so called Curry-Howard-Lawvere/Lambek correspondence by adding category theory as shown in the following table.

@@

Types consists of ground types $G$, the unit type $\unit$, product types $A \times B$ and function types $A \to B$.
Possible ground types are \textsf{int} and \textsf{bool}.
Terms are variables $x$, constants $c^A$ annotated by its type $A$, the empty tuple $()$, pairs $(s, t)$, first ($\fst s$) and second ($\snd s$) projections,
function definitions $\lambda x: A.\, s$ and function applications $s\,t$.
Common constants are integers like $3^{\mathsf{int}}$ and booleans ($\mathsf{true}^{\mathsf{bool}}$ and $\mathsf{false}^{\mathsf{bool}}$).
In order to state that a term has some type we use environments.
Environments are lists of variables and their types.
Thus we have the empty environment $\eenv$ and we can extend an environment $\Gamma$ with a variable $x$ of type $A$ (written $\Gamma, x: A$).
Often the empty environment is ommited.
For example the environment $x: \mathsf{int}, y: \mathsf{bool}$ has a variable $x$ of type \textsf{int} and a variable $y$ of type \textsf{bool}.
Variables in environments are always distinct, that means $x: \mathsf{int}, x: \mathsf{int}$ is not an environment.
We use environments to `remember' the type of a variable in a function definition.
The following definition shows how environments $\Gamma$ are used to state that a term $s$ has a type $A$.
We write $\Gamma \type s: A$ for this statement.
If the environment is empty we write just $\type s: A$ instead of $\eenv \type s: A$.
Also, we say that a term $s$ is well-typed, if an  environment $\Gamma$ and a type $A$ exists, such that the judgement $\Gamma \type s: A$ holds.

@@

\begin{defn}
    \label{def:ccc}
    A cartesian closed category $\cat{C}$ consists of:
    \begin{itemize}
        \item A teminal object $\term \in \cat{C}$: $\forall X \in \cat{C}.\, \exists! \langle\rangle: X \to \term$.
        \item Binary products $X \times Y \in \cat{C}$ with projections $\pi_1 \from X \times Y \to X$ and $\pi_2 \from X \times Y \to Y$
            for all $X, Y \in \cat{C}$:
            For every $Z \in \cat{C}$ and morphisms $f \from Z \to X$ and $g \from Z \to Y$,
            there is exactly one morphism $\langle f, g \rangle \from Z \to X \times Y$, such that the following diagram commutes:
            \[
                \begin{tikzcd}
                    & Z \arrow[bend right]{ddl}{f} \arrow{d}{\langle f, g \rangle} \arrow[bend left]{ddr}{g} & \\
                    & X \times Y \arrow{dl}{\pi_1} \arrow{dr}{\pi_2} & \\
                    X & & Y
                \end{tikzcd}
            \]
            For two morphisms $p \from X \to X'$ and $q \from Y \to Y'$ we define
            \[ p \times q := \langle p \circ \pi_1, q \circ \pi_2 \rangle \from X \times Y \to X' \times Y'.\]
        \item Exponentials $Y^X \in \cat{C}$ together with a morphism $\app \from Y^X \times X \to Y$ for all $X, Y \in \cat{C}$:
            For every $Z \in \cat{C}$ and morphism $f \from Z \times X \to Y$,
            there is a unique morphism $\cur(f) \from Z \to Y^X$, such that following diagram commutes:
            \[
                \begin{tikzcd}
                    Z \times X \arrow{d}[left]{\cur(f) \times \id_X} \arrow{r}{f} & Y \\
                    Y^X \times X \arrow{ur}[right=7pt, below]{\app}
                \end{tikzcd}
            \]
    \end{itemize}
\end{defn}

@@

%The case for constants 
\todo{explain, esp. $M(c^A)$}

@@

In the previous section we saw how to express well-typed terms as morphisms in a cartesian closed category.
Example~\ref{exp:sem} showed that there are terms which are mapped to the same morphism.
It turns out that this holds for all $\beta$- and $\eta$-equivalent terms.
In order to define this equivalence we need substitutions.
A substitution corresponds to a function application $(\lambda x: A.\, s)\, t$, where the function is called with an argument $t$.
It replaces in the term $s$ all occurences of the variable $x$ with the term $t$.
We write $s[t/x]$ and define it as follows.

>>>>

Recall the Curry-Howard correspondence which relates logic and type theory.
According to this correspondence propositions are types and proofs are terms.
A prominent example is the relation between intuitionistic propositional logic and simply typed lambda calculus.
We can extend the correspondence to the so called Curry-Howard-Lawvere/Lambek correspondence by adding category theory as shown in the following table.

@@

Types consists of ground types $G$, the unit type $\unit$, product types $A \times B$ and function types $A \to B$.
Possible ground types are \textsf{int} and \textsf{bool}.
Terms are variables $x$, constants $c^A$ annotated by its type $A$, the empty tuple $()$, pairs $(s, t)$, first ($\fst s$) and second ($\snd s$) projections,
function definitions $\lambda x: A.\, s$ and function applications $s\,t$.
Common constants are integers like $3^{\mathsf{int}}$ and booleans ($\mathsf{true}^{\mathsf{bool}}$ and $\mathsf{false}^{\mathsf{bool}}$).
In order to state that a term has some type we use environments.
Environments are lists of variables and their types.
Thus we have the empty environment $\eenv$ and we can extend an environment $\Gamma$ with a variable $x$ of type $A$ (written $\Gamma, x: A$).
Often the empty environment is omitted.
For example the environment $x: \mathsf{int}, y: \mathsf{bool}$ has a variable $x$ of type \textsf{int} and a variable $y$ of type \textsf{bool}.
Variables in environments are always distinct, that means $x: \mathsf{int}, x: \mathsf{int}$ is not an environment.
We use environments to `remember' the type of a variable in a function definition.
The following definition shows how environments $\Gamma$ are used to state that a term $s$ has a type $A$.
We write $\Gamma \type s: A$ for this statement.
If the environment is empty we write just $\type s: A$ instead of $\eenv \type s: A$.
Also, we say that a term $s$ is well-typed, if an  environment $\Gamma$ and a type $A$ exists, such that the judgement $\Gamma \type s: A$ holds.

@@

\begin{defn}
    \label{def:ccc}
    A cartesian closed category $\cat{C}$ consists of:
    \begin{itemize}
        \item A terminal object $\term \in \cat{C}$: $\forall X \in \cat{C}.\, \exists! \langle\rangle: X \to \term$.
        \item Binary products $X \times Y \in \cat{C}$ with projections $\pi_1 \from X \times Y \to X$ and $\pi_2 \from X \times Y \to Y$
            for all $X, Y \in \cat{C}$:
            For every $Z \in \cat{C}$ and morphisms $f \from Z \to X$ and $g \from Z \to Y$,
            there is exactly one morphism $\langle f, g \rangle \from Z \to X \times Y$, such that the following diagram commutes:
            \[
                \begin{tikzcd}
                    & Z \arrow[bend right]{ddl}{f} \arrow{d}{\langle f, g \rangle} \arrow[bend left]{ddr}{g} & \\
                    & X \times Y \arrow{dl}{\pi_1} \arrow{dr}{\pi_2} & \\
                    X & & Y
                \end{tikzcd}
            \]
            For two morphisms $p \from X \to X'$ and $q \from Y \to Y'$ we define
            \[ p \times q := \langle p \circ \pi_1, q \circ \pi_2 \rangle \from X \times Y \to X' \times Y'.\]
        \item Exponentials $Y^X \in \cat{C}$ together with a morphism $\app \from Y^X \times X \to Y$ for all $X, Y \in \cat{C}$:
            For every $Z \in \cat{C}$ and morphism $f \from Z \times X \to Y$,
            there is a unique morphism $\cur(f) \from Z \to Y^X$, such that following diagram commutes:
            \[
                \begin{tikzcd}
                    Z \times X \arrow{d}[left]{\cur(f) \times \id_X} \arrow{r}{f} & Y \\
                    Y^X \times X \arrow{ur}[right=7pt, below]{\app}
                \end{tikzcd}
            \]
    \end{itemize}
\end{defn}

@@

The case for constants illustrates why we choose $M(c^A) \from \term \to M|[A|]$.
Since the type of a constant does not depend on the environment $\Gamma$, it is natural to take the empty environment.
We can easily construct morphisms for arbitrary environments, because the empty environment is mapped to the terminal object.

@@

In the previous section we saw how to express well-typed terms as morphisms in a cartesian closed category.
Example~\ref{exp:sem} showed that there are terms which are mapped to the same morphism.
It turns out that this holds for all $\beta$- and $\eta$-equivalent terms.
In order to define this equivalence we need substitutions.
A substitution corresponds to a function application $(\lambda x: A.\, s)\, t$, where the function is called with an argument $t$.
It replaces in the term $s$ all occurrences of the variable $x$ with the term $t$.
We write $s[t/x]$ and define it as follows.
